\vskip-0.6in Trade Agreements in the Shadow of Lobbying\protect\footnote{Formerly circulated under the title ``Trade Agreements, Lobbying and Separation of Powers''}

### 2016-02-02 draft.tex
revise abstract, end of intro on PFS puzzle

\begin{quote}
{This paper presents a model of international trade agreements in which the executive branches of each government negotiate agreements while the legislative branches, subject to political pressure from firms, can disrupt them. Lobbying is in the style of Grossman and Helpman (1994) with a new feature: all actors face uncertainty arising from the complexity of the legislative process. I demonstrate that the lower the executives set tariffs in a trade agreement, the more effort lobbies put forth to prevent its ratification. Thus trade agreements act as a domestic political commitment device: executives set relatively high tariffs to discourage lobbying and increase the chance that the agreement will be ratified. This reconciles the result from tests of Grossman and Helpman's model that protection levels are high relative to contributions given estimates of governments' social-welfare weights.

\textit{JEL classification:} D72, D80, F13, F53, F55 \\
\textit{Keywords:} trade policy, lobbying, political economy, protection, international trade agreements, uncertainty}
\end{quote}

@@

Assuming for now that lobbies represent only import-competing interests, lobbies in this environment would like to prevent or derail a trade agreement that would reduce their tariff protection. They work toward this end by exerting effort to convince their legislatures to vote against the trade agreement but face some uncertainty about whether their efforts will result in a voting outcome consistent with their interests. I show that the probability that the trade agreement is derailed increases in lobbying effort and decreases in the trade agreement tariff. Further, the lobby decreases its effort as the tariff it would face under the trade agreement rises: it has less to lose the higher is the trade agreement tariff. When forming the trade agreement tariff, the executives trade off the welfare they receive under the agreement with the probability that the agreement actually remains in force. Ultimately, I find that observed lobbying effort may often be low because the executive branches of the governments set trade agreement tariffs sufficiently high to reduce political pressure on the legislatures.

@@

Ultimately, it is the addition of a rich structure of government, tailored carefully to the trade policy making process, that resolves the main empirical puzzle surrounding the PFS model: in this framework, even a welfare-maximizing executive and lobbying effort at zero can be consistent with high tariff levels.

\vskip.2in
I find that the separation-of-powers government structure can explain the empirical puzzle surrounding governments' welfare weights and that the addition of political uncertainty adds important realism so that the model presented here can speak to important trade policy-making events such as the four-year delay in the ratification of the U.S.-Korea Free Trade Agreement.

Empirical investigations of Grossman and Helpman's (1994) PFS model broadly support the predictions of the theory qualitatively, yet their quantitative estimates present several interesting puzzles.\footnote{cfr. Goldberg and Maggi (1999); Gawande and Bandyopadhyay (2000); Mitra, Thomakos and Ulubasoglu (2002); McCalman (2004).\label{fn:gh_lobby}} These studies have consistently found the weight governments place on social welfare to be many times that which they place on lobbying effort, while estimates indicate that the deadweight losses caused by trade distortions are several orders of magnitude larger than lobbying expenditures.\footnote{Feenstra (1992) assembles estimates from the mid-1980s that place a floor of $\$$8 billion per year on the deadweight losses from protection, whereas Bombardini and Trebbi (2012) find that total annual lobbying expenditure on trade issues in 1999-2001 when this data first became available was about $\$$200 million.} This raises the question: how could governments value social welfare so highly, yet grant these quantities of protection at such a low price?

@@

In the non-unitary government model, these estimates and the earlier-cited stylized facts become consistent with each other. The ``case of the missing contributions,'' as Gawande and Hoekman (2006) refer to it, turns out to be well explained by an equilibrium phenomenon arising from the agenda-setting power of the executive branch. In particular, this model can explain the approximately 15$\%$ of sectors that receive protection in spite of apparently putting forth no lobbying effort.

>>>>

\begin{quote}
{This paper presents a model of international trade agreements in which the executive branches of each government negotiate agreements while the legislative branches, subject to political pressure from firms, can disrupt them. Lobbying is in the style of Grossman and Helpman (1994) with a new feature: all actors face uncertainty arising from the complexity of the legislative process. I demonstrate that the lower the executives set tariffs in a trade agreement, the more effort lobbies put forth to prevent its ratification. Thus trade agreements act as a domestic political commitment device: executives set relatively high tariffs to discourage lobbying and increase the chance that the agreement will be ratified. This finding sheds light on the empirical puzzle surrounding governments' welfare weights in the Grossman and Helpman (1994) model, while the addition of political uncertainty adds important realism so that the model presented here can speak to important trade policy-making events such as the four-year delay in the ratification of the U.S.-Korea Free Trade Agreement.
\textit{JEL classification:} D72, D80, F13, F53, F55 \\
\textit{Keywords:} trade policy, lobbying, political economy, protection, international trade agreements, uncertainty}
\end{quote}

@@

Assuming for now that lobbies represent only import-competing interests, lobbies in this environment would like to prevent or derail a trade agreement that would reduce their tariff protection. They work toward this end by exerting effort to convince their legislatures to vote against the trade agreement but face some uncertainty about whether their efforts will result in a voting outcome consistent with their interests. I show that the probability that the trade agreement is derailed increases in lobbying effort and decreases in the trade agreement tariff. Further, the lobby decreases its effort as the tariff it would face under the trade agreement rises: it has less to lose the higher is the trade agreement tariff. When forming the trade agreement tariff, the executives trade off the welfare they receive under the agreement with the probability that the agreement actually remains in force. Ultimately, I find that observed lobbying effort may often be low because the executive branches of the governments set trade agreement tariffs sufficiently high to reduce political pressure on the legislatures. I also show that other examples of realism are gained when political uncertainty is added to the model: for instance, the model rationalizes the four-year delay and eventual ratification of the U.S.-Korea Free Trade Agreement.

@@

The addition of a rich structure of government, tailored carefully to the trade policy making process, helps shed light on the empirical puzzle surrounding the PFS model. Empirical tests of the model\footnote{cfr. Goldberg and Maggi (1999); Gawande and Bandyopadhyay (2000); Mitra, Thomakos and Ulubasoglu (2002); McCalman (2004).\label{fn:gh_lobby}} have consistently found the weight governments place on social welfare to be many times that which they place on lobbying effort, while estimates indicate that the deadweight losses caused by trade distortions are several orders of magnitude larger than lobbying expenditures.\footnote{Feenstra (1992) assembles estimates from the mid-1980s that place a floor of $\$$8 billion per year on the deadweight losses from protection, whereas Bombardini and Trebbi (2012) find that total annual lobbying expenditure on trade issues in 1999-2001 when this data first became available was about $\$$200 million.} This raises the question: how could governments value social welfare so highly, yet grant these quantities of protection at such a low price?

@@

In the non-unitary government model, these estimates and the earlier-cited stylized facts become consistent with each other. The ``case of the missing contributions,'' as Gawande and Hoekman (2006) refer to it, turns out to be well explained by an equilibrium phenomenon arising from the agenda-setting power of the executive branch. In particular, in this model, there is inconsistency for the approximately 15$\%$ of sectors that receive protection but apparently put forth no lobbying effort. In this framework, even a welfare-maximizing executive and lobbying effort at zero can be reconciled with high tariff levels.
